---
title: "Finding the optimal capacity mix"
author: "Uni Lee"
date: "3/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(stringr)
```

# How does renewable energy change the optimal capacity mix for electricity generation? 

As Energiewende (energy transition) advances around the world, the electricity sector is going through a major structural change. Countries around the world are moving away from fossil fuels and nuclear power and increasing the share of low-carbon energy sources to reduce carbon emissions. How would the higher share of renewable energy change the optimal capacity mix? To answer this question, let us turn to Denmark, [which has the largest share of low-carbon energy (solar, wind, hydro, biomass, waste, geothermal) in the EU, reaching 77.4% of the total generation in 2020.  (data source: Our World in Data)](https://github.com/owid/energy-data). 

```{r}
eu <- c("AUT", "BEL", "BGR", "HRV", "CYP", "CZE", "DNK", "EST", "FIN", "FRA", "DEU", "GRC", "HUN", "IRL", "ITA", "LVA", "LTU", "LUX", "MLT", "NLD", "POL", "PRT", "ROU", "SVK", "SVN", "ESP", "SWE")
energy <- read.csv("data/owid-energy-data.csv", sep=",") %>%
  filter(year==2019, iso_code %in% eu) %>% select(country, renewables_share_elec) %>% na.omit() %>%
  arrange(desc(renewables_share_elec)) %>% top_n(10)

ggplot(energy, aes(x=reorder(country,-renewables_share_elec), y=renewables_share_elec)) + geom_col() +
  ylim(0,90)

```


## Data

[Open Power System Data](https://open-power-system-data.org/) gives hour-by-hour electricity demand (load), total capacity, total generation, solar and wind generation and capacity in the EU. 

For this exercise, we take a look at 

```{r}
raw <- read.csv("data/openpowersystemdata.csv")
denmark <- raw %>% filter(str_detect(cet_cest_timestamp, "2019")) %>%  
  select(timestamp=cet_cest_timestamp,
         load = DK_load_actual_entsoe_transparency,
         solar_gen = DK_solar_generation_actual,
         wind_gen = DK_wind_generation_actual) %>%
  mutate(residual_load = load - solar_gen - wind_gen)  # Calculate residual load, the rest of the load not served by solar and wind

# Order the residual load in descending order 
denmark_arr <- denmark %>% 
  arrange(desc(residual_load)) 

# Create a vector to name each hour of the year
hr <- 1:8760 

denmark_arr <- cbind(hr, denmark_arr)

# Plot the load duration curve 
ggplot(denmark_arr, aes(x=hr, y=residual_load)) + geom_area()
```

# Screening Curve



